Evaluation of probabilistic photometric redshift estimation approaches for \textsc{LSST}

### 2019-03-15 paper/tex/analysis.tex
finished statistics

\subsection{Metrics of an ensemble of photo-\mathinhead{$z$}{} posteriors}
\label{sec:qualmet}
%(Rongpu Zhou, Jeffrey Newman, Alex Malz)
%(Alex Malz, Peter Freeman, Sam Schmidt)

Because \lsst's \pzpdf s will be used for many scientific applications, some of which require accuracy of each individual catalog entry, we consider several metrics that probe the population-level performance of the \pzpdf s.
Because we have the true redshifts but not true \pzpdf s for comparison, we remind the reader of the Cumulative Distribution Function (CDF)
\begin{equation}
  \label{eq:cdf}
  \rm{CDF}(q) \equiv \int_{-\infty}^{q} p(z) dz,
\end{equation}
of a \pzpdf, which is used as the basis for several of our metrics.

@@

The probability integral transform (PIT)
\begin{align}
\label{eq:pit}
\mathrm{PIT} &\equiv \int_{-\infty}^{z_{\rm true}} p(z) dz
\end{align}
is the CDF of a \pzpdf\ evaluated at its true redshift, and the distribution of PIT values probes the average accuracy of the \pzpdf s of an ensemble of galaxies.
The distribution of PIT values is effectively the derivative of the QQ plot.
A catalogue of accurate \pzpdf s should have a PIT distribution that is uniform $U(0,1)$, and deviations from flatness are interpretable: overly broad \pzpdf s induce underrepresentation of the lowest and highest PIT values, whereas overly narrow \pzpdf s induce overrepresentation of the lowest and highest PIT values.
Catastrophic outliers with a true redshift outside the support of its \pzpdf\ have $\mathrm{PIT} \approx 0$ or $\mathrm{PIT} \approx 1$.

The PIT distribution has been used to quantify the performance of \pzpdf\ methods in the past \citep[e.~g.~][]{Bordoloi:10,Polsterer:16,Tanaka:17}.
\citet{Tanaka:17} use the histogram of PIT values as a diagnostic indicator of overall code performance, while \citet{Freeman:17} independently define the PIT and demonstrate how its individual values may be used both to perform hypothesis testing (via, e.g., the KS, CvM, and AD tests; see below) and to construct quantile-quantile plots.
%\red{We might want to divide this by redshift to get trends, as the overall PIT could be impacted by true n(z).
%For example, a PIT with nontrivial shape could be due to that behavior being uniform as a function of redshift or significant inaccuracy only in a small range of redshifts where $n(z)$ is large and a flat PIT elsewhere.}
Though the QQ plot contains very similar information to that shown in the PIT histogram plot, due to being the PIT being the derivative of the QQ, we include both forms for completeness and enhanced visual interpretability.

\subsection{Conditional density estimation loss}
\label{sec:CDE_loss}
%(Alex Malz, Ann Lee)

With the conditional density estimation loss (CDE loss) we can compare how well different methods estimate individual PDFs for photometric covariates $\x$ rather than looking only at the ensemble distribution.
As in Section~\ref{sec:flexzboost}, we use the notation $f(z \vert \x)$ instead of $p(z)$ to explicitly show the dependence on the photometry $\x$.

%\aim{Perhaps this should be rewritten in the same notation used elsewhere in the paper.} \red{I agree, does someone who speaks both stats and astro want to take a shot at this?}
%\textcolor{cyan}{Ann: We are looking at individual PDFs so the dependence on $\x$ needs to be explicit; this is probably the most accurate notation as it shows both the conditioning clearly as well as whether $\x$ is fixed or random.}

The CDE loss is defined as
\begin{equation} \label{eq:cde-loss}
L(f, \widehat{f}) \equiv \int \int (f(z \mid \x) - \widehat{f}(z \mid \x))^{2} dz dP(\x) .
\end{equation}

This loss is the CDE equivalent of the RMSE in regression.
To estimate this loss we rewrite it as
\begin{equation} \label{eq:estimated-cde-loss}
L(f, \widehat{f}) = \mathbb{E}_\X \left[\int \widehat{f}(z \mid \X)^{2} dz\right] - 2 \mathbb{E}_{\X,Z}\left[\widehat{f}(Z \mid \X)\right] + K_{f},
\end{equation}
where upper-case letters denote random variables and lower-case the observed variables.  The first expectation is with respect to the marginal distribution of the covariates $\X$, the second expectation is with respect to the joint distribution of $\X$ and $Z$, and $K_{f}$ is a constant depending only upon the true conditional densities $f(z\mid \x)$.
For each method we can estimate these expectations as empirical expectations on the test or validation data \citep[Eq.~7 in][]{Izbicki:17b} without knowledge of the true densities.

@@

\subsection{Metrics over estimated probability distributions}
\label{sec:quantmet}

In tandem with the QQ and PIT metrics introduced above, we additionally compute the following metrics comparing the empirical CDF of a distribution to the true or expected distribution.
These metrics give a more quantitative measure of the departure from ideal than the more visual PIT histogram and QQ plot.
We compute metrics comparing the CDF of PIT values to the CDF of a Uniform distribution, and also compute the CDF of the true redshift distribution $N'(z)$ compared the $\hat{N}(z)$ distribution derived from summing the \pzpdf s as described in Eq.~\ref{eq:stacked}.

%\subsubsection{Root-mean-square error (RMSE)}
%\label{sec:rmse}
%%(Alex Malz)%
%
%We employ the familiar root-mean-square error
%\begin{equation}
%\mathrm{RMSE} \equiv \sqrt{\int_{-\infty}^{\infty}\left(\hat{f}(z)-f'(z)\right)^{2}\,dz} .
%\end{equation}
%%may be used for two purposes in this analysis.
%%We calculate it between the true redshift distribution $N'(z)$ and each estimator $\hat{N}(z)$ (derived by Eq.~\ref{eq:stacked}) as well as between the ideal QQ and PIT curves from each method studied and the ideal curves.
%Though this metric does not account for the fact that the redshift distribution function is, in fact, a probability distribution, it can still be interpreted as a measure of the integrated difference between the estimated distribution and the true distribution.%, and it can be used to quantify the otherwise qualitative metrics.
%

\subsubsection{Kolmogorov-Smirnov (KS) and related statistics}
\label{sec:ks}
%(John Soo, Rongpu Zhou)

The \textit{Kolmogorov-Smirnov statistic} $N_{\rm KS}$ is the maximum difference between $F_{\rm phot}(z)$ and $F_{\rm spec}(z)$, the CDFs of the photo-$z$ and spectroscopic redshift respectively:
\begin{equation}
N_{\rm KS} \equiv \max_{z}\left( \left| F_{\rm phot}(z) - F_{\rm spec}(z) \right| \right).
\end{equation}
%The KS test quantifies whether the two redshift distributions are drawn from a similar distribution, independent of binning.
The KS test quantifies the similarity between two distributions, independent of binning.
A lower $N_{\rm KS}$ value corresponds to more similar distributions.
%We calculate two KS test statistics, one to test the distribution of $p(z)$ values, and one to test the stacked estimator $\hat{n}(z)$ of the redshift distribution.
%The $p(z)$ test compares the CDF of the PIT values described in the previous section to that of a uniform distribution.
%The $n(z)$ test compares the CDF of the distribution formed by stacking the $p(z)$ for every object in our test set against the CDF of the corresponding true redshifts.
%In this case a lower $N_{\rm KS}$ value means that the photo-$z$ distribution fits the spectroscopic redshift distribution well. $F_{\rm phot}(z)$ is calculated by stacking the PDFs of every single object in the sample, producing a smooth cumulative $n(z)$ to be compared with $F_{\rm spec}(z)$.

We also consider two variants of the KS statistic: the Cramer-von Mises (CvM) and Anderson-Darling (AD) statistics.
The CvM statistic is similar to the KS statistic as it is also computed from the distance between the measured CDF and the ideal CDF, but instead of the maximum distance, the CvM statistic
\begin{equation}
\label{eq:cvm}
\omega^2 \equiv \int_{-\infty}^{+\infty}\big(F_{\mathrm{meas.}}(x) - F_{\mathrm{ideal}}(x)\big)^2\mathrm{d}F_{\mathrm{ideal}}
\end{equation}
is the average of the distance squared.
% In this paper we use the CvM statistic to evaluate the performance of both $n(z)$ and $p(z)$.

The AD statistic
\begin{equation} \label{eq:ad}
A^2 \equiv N_{tot}\int_{-\infty}^{+\infty} \frac{\big(F_{\mathrm{meas.}}(x) - F_{\mathrm{ideal}}(x)\big)^2} {F_{\mathrm{ideal}}(x) (1-F_{\mathrm{ideal}}(x))}\mathrm{d}F_{\mathrm{ideal}}
\end{equation}
is a weighted version of the CvM statistic, making it more sensitive to the tails of the distribution, where $N_{tot}$ is the sample size.
%The KS statistic and the CvM and AD statistics related to it are all used to evaluate the performance of both $\hat{n}(z)$ and the QQ and PIT of the CDFs of the ensemble of photo-$z$ interim posteriors.

@@

The most popular application of \pzpdf s is the estimation of the overall redshift distribution $N(z)$, the true value of which is known for the DC1 data set and will be denoted as $\tilde{N}(z)$.
In terms of the prior information provided to each method, the true redshift distribution satisfies the tautology $\tilde{N}(z) = p(z \vert I_{D})$ due to our experimental set-up; because the DC1 training and template sets are representative and complete, $I_{D}$ represents a prior that is also equal to the truth.
In this ideal case of complete and representative prior information, the method that would give the best approximation to $\hat{N}(z)$ would be one that neglects all the information contained in the photometry $\{d_{i}\}_{N_{tot}}$ and gives every galaxy the same \pzpdf\ $\hat{p}_{i}(z) = \tilde{N}(z)$ for all $i$; the inclusion of any information from the photometry would only introduce noise to the optimal result of returning the prior.
This is the exact estimator, \trainz, that we have described in Section~\ref{sec:method:trainz}, and which will serve as an experimental control.

@@

We use the QQ plot to visualize the relationship between $\hat{N}^{H}(z)$ and $\tilde{N}(z)$.
We additionaly calculate the first three moments of the estimated redshift distribution $\hat{N}^{H}(z)$ for each code and compare them to the moments of the true redshift distribution $N'(z)$.
The $m^{\mathrm{th}}$ moment of a distribution is defined as
\begin{equation}
\langle z^{m}\rangle \equiv \int_{-\infty}^{\infty}z^{m}N(z)dz .
\end{equation}
Here, we use the moments of the stacked estimator of the redshift distribution function as the basis for a metric.
The closer the moments of $\hat{N}(z)$ for a \pzpdf\ method are to the moments of the true redshift distribution function $N'(z)$, the better the \pzpdf\ method is a estimating the overall redshift distribution ``shape''.

>>>>

\subsection{Metric of individual photo-\mathinhead{$z$}{} posteriors}
\label{sec:CDE_loss}
%(Alex Malz, Ann Lee)

To the knowledge of the authors, there is only one metric that can be used to evaluate the performance of individual \pzpdf\ estimators in the absence of true \pz\ posteriors.
Using the notation introduced in Section~\ref{sec:flexzboost}, the conditional density estimation (CDE) loss is defined as
\begin{equation}
  \label{eq:cde-loss}
  L(f, \widehat{f}) \equiv \int \int (f(z \vert \x) - \widehat{f}(z \vert \x))^{2} dz dP(\x)
\end{equation}
in terms of the photometry $\x$, an analogue to the familiar root-mean-square-error used in conventional regression.
We estimate the CDE loss via
\begin{equation}
  \label{eq:estimated-cde-loss}
  \hat{L}(f, \widehat{f}) = \mathbb{E}_\X \left[\int \widehat{f}(z \mid \X)^{2} dz\right] - 2 \mathbb{E}_{\X, Z}\left[\widehat{f}(Z \mid \X)\right] + K_{f},
\end{equation}
where the first term is the expectation value of the \pz\ posterior with respect to the marginal distribution of the photometric covariates $\X$, the second term is the expectation value with respect to the joint distribution of $\X$ and the space $Z$ of all possible redshifts, and the third term $K_{f}$ is a constant depending only upon the true conditional densities $f(z \vert \x)$.
We may estimate these expectations empirically on the test or validation data \citep[Eq.~7 in][]{Izbicki:17b} without knowledge of the true densities.

\subsection{Metrics of an ensemble of photo-\mathinhead{$z$}{} posteriors}
\label{sec:qualmet}
%(Rongpu Zhou, Jeffrey Newman, Alex Malz)
%(Alex Malz, Peter Freeman, Sam Schmidt)
%(John Soo, Rongpu Zhou)

Because \lsst's \pzpdf s will be used for many scientific applications, some of which require accuracy of each individual catalog entry, we consider several metrics that probe the population-level performance of the \pzpdf s.
Because we have the true redshifts but not true \pzpdf s for comparison, we remind the reader of the Cumulative Distribution Function (CDF)
\begin{equation}
  \label{eq:cdf}
  \rm{CDF}[f, q] \equiv \int_{-\infty}^{q} f(z) dz,
\end{equation}
of a generic univariate PDF $f(z)$, which is used as the basis for several of our metrics.

@@

The probability integral transform (PIT)
\begin{align}
\label{eq:pit}
\mathrm{PIT} &\equiv \rm{CDF}[\hat{p}, z_{\rm true}] %\int_{-\infty}^{z_{\rm true}} p(z) dz
\end{align}
is the CDF of a \pzpdf\ evaluated at its true redshift, and the distribution of PIT values probes the average accuracy of the \pzpdf s of an ensemble of galaxies.
The distribution of PIT values is effectively the derivative of the QQ plot.
A catalogue of accurate \pzpdf s should have a PIT distribution that is uniform $U(0,1)$, and deviations from flatness are interpretable: overly broad \pzpdf s induce underrepresentation of the lowest and highest PIT values, whereas overly narrow \pzpdf s induce overrepresentation of the lowest and highest PIT values.
Catastrophic outliers with a true redshift outside the support of its \pzpdf\ have $\mathrm{PIT} \approx 0$ or $\mathrm{PIT} \approx 1$.

The PIT distribution has been used to quantify the performance of \pzpdf\ methods in the past \citep[e.~g.~][]{Bordoloi:10,Polsterer:16,Tanaka:17}.
\citet{Tanaka:17} use the histogram of PIT values as a diagnostic indicator of overall code performance, while \citet{Freeman:17} independently define the PIT and demonstrate how its individual values may be used both to perform hypothesis testing (via, e.g., the KS, CvM, and AD tests; see below) and to construct quantile-quantile plots.
%\red{We might want to divide this by redshift to get trends, as the overall PIT could be impacted by true n(z).
%For example, a PIT with nontrivial shape could be due to that behavior being uniform as a function of redshift or significant inaccuracy only in a small range of redshifts where $n(z)$ is large and a flat PIT elsewhere.}

@@

% \subsection{Metrics over estimated probability distributions}
% \label{sec:quantmet}

% We compute metrics comparing the CDF of PIT values to the CDF of a Uniform distribution, and also compute the CDF of the true redshift distribution $N'(z)$ compared the $\hat{N}(z)$ distribution derived from summing the \pzpdf s as described in Eq.~\ref{eq:stacked}.

%\subsubsection{Root-mean-square error (RMSE)}
%\label{sec:rmse}
%%(Alex Malz)%
%We employ the familiar root-mean-square error
%\begin{equation}
%\mathrm{RMSE} \equiv \sqrt{\int_{-\infty}^{\infty}\left(\hat{f}(z)-f'(z)\right)^{2}\,dz} .
%\end{equation}
%%may be used for two purposes in this analysis.
%%We calculate it between the true redshift distribution $N'(z)$ and each estimator $\hat{N}(z)$ (derived by Eq.~\ref{eq:stacked}) as well as between the ideal QQ and PIT curves from each method studied and the ideal curves.
%Though this metric does not account for the fact that the redshift distribution function is, in fact, a probability distribution, it can still be interpreted as a measure of the integrated difference between the estimated distribution and the true distribution.%, and it can be used to quantify the otherwise qualitative metrics.

% \subsubsection{Kolmogorov-Smirnov (KS) and related statistics}
% \label{sec:ks}

We evaluate a number of quantitative metrics derived from the visually interpretable QQ plot and PIT histogram, built on the Kolmogorov-Smirnov (KS) statistic
\begin{equation}
  \label{eq:ks}
  \rm{KS} \equiv \max_{z} \left( \left| \rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z] \right| \right),
\end{equation}
interpretable as the maximum difference between the CDFs of an approximating univariate distribution $\hat{f}(z)$ and a reference distribution $\tilde{f}(z)$.
% $CDF_{\rm phot}(z)$ and $CDF_{\rm spec}(z)$, the CDFs of the photo-$z$ and spectroscopic redshift respectively:
%The KS test quantifies whether the two redshift distributions are drawn from a similar distribution, independent of binning.
% The KS test quantifies the similarity between two distributions, independent of binning.
% A lower $N_{\rm KS}$ value corresponds to more similar distributions.
%We calculate two KS test statistics, one to test the distribution of $p(z)$ values, and one to test the stacked estimator $\hat{n}(z)$ of the redshift distribution.
%The $p(z)$ test compares the CDF of the PIT values described in the previous section to that of a uniform distribution.
%The $n(z)$ test compares the CDF of the distribution formed by stacking the $p(z)$ for every object in our test set against the CDF of the corresponding true redshifts.
%In this case a lower $N_{\rm KS}$ value means that the photo-$z$ distribution fits the spectroscopic redshift distribution well. $F_{\rm phot}(z)$ is calculated by stacking the PDFs of every single object in the sample, producing a smooth cumulative $n(z)$ to be compared with $F_{\rm spec}(z)$.
We also consider two variants of the KS statistic.
A cousin of the KS statistic, the Cramer-von Mises (CvM) statistic
\begin{equation}
\label{eq:cvm}
  \rm{CvM}^{2} \equiv \int_{-\infty}^{+\infty} \big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2 \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is the mean-squared difference between the CDFs of an approximate and true PDF.
The Anderson-Darling (AD) statistic
\begin{equation} \label{eq:ad}
  \rm{AD}^2 \equiv N_{tot}\int_{-\infty}^{+\infty} \frac{\big(\rm{CDF}[\hat{f}, z] - \rm{CDF}[\tilde{f}, z]\big)^2} {\rm{CDF}[\tilde{f}, z] (1 -\rm{CDF}[\tilde{f}, z]} \mathrm{d}\rm{CDF}[\tilde{f}, z]
\end{equation}
is a weighted mean-squared difference featuring enhanced sensitivity to discrepancies in the tails of the distribution.
% The CvM statistic is similar to the KS statistic as it is also computed from the distance between the measured CDF and the ideal CDF, but instead of the maximum distance, the CvM statistic
% is the average of the distance squared.
% In this paper we use the CvM statistic to evaluate the performance of both $n(z)$ and $p(z)$.
% The AD statistic
% is a weighted version of the CvM statistic, making it more sensitive to the tails of the distribution, where $N_{tot}$ is the sample size.
%The KS statistic and the CvM and AD statistics related to it are all used to evaluate the performance of both $\hat{n}(z)$ and the QQ and PIT of the CDFs of the ensemble of photo-$z$ interim posteriors.

@@

Perhaps the most popular application of \pzpdf s is the estimation of the overall redshift distribution $N(z)$, a quantity that enters some cosmological calculations and the true value of which is known for the DC1 data set and will be denoted as $\tilde{N}(z)$.
In terms of the prior information provided to each method, the true redshift distribution satisfies the tautology $\tilde{N}(z) = p(z \vert I_{D})$ due to our experimental set-up; because the DC1 training and template sets are representative and complete, $I_{D}$ represents a prior that is also equal to the truth.
In this ideal case of complete and representative prior information, the method that would give the best approximation to $\hat{N}(z)$ would be one that neglects all the information contained in the photometry $\{d_{i}\}_{N_{tot}}$ and gives every galaxy the same \pzpdf\ $\hat{p}_{i}(z) = \tilde{N}(z)$ for all $i$; the inclusion of any information from the photometry would only introduce noise to the optimal result of returning the prior.
This is the exact estimator, \trainz, that we have described in Section~\ref{sec:method:trainz}, and which will serve as an experimental control.

@@

As $N(z)$ is itself a univariate PDF, we apply the metrics of the previous sections to it as well.
We additionaly calculate the first three moments
\begin{equation}
  \label{eq:moment}
\langle z^{m}\rangle \equiv \int_{-\infty}^{\infty} z^{m} N(z) dz
\end{equation}
of the estimated redshift distribution $\hat{N}^{H}(z)$ for each code and compare them to the moments of the true redshift distribution $\tilde{N}(z)$.
Under the assumption that the stacked estimator is unbiased, a superior method minimizes the difference between the true and estimated moments.
